\documentclass[acmsmall,screen,review,anonymous,nonacm]{acmart}

\let\Bbbk\relax
\usepackage{amsmath,amssymb}
\usepackage{amsthm}
\newtheorem{remark}{Remark}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,calc}

%% Lean listing style
\definecolor{leankeyword}{RGB}{0,0,255}
\definecolor{leancomment}{RGB}{0,128,0}
\definecolor{leanstring}{RGB}{163,21,21}
\lstdefinelanguage{Lean}{
  morekeywords={theorem,lemma,def,inductive,structure,where,import,open,namespace,end,section,variable,by,have,exact,rcases,intro,apply,sorry,noncomputable,instance,class,abbrev,Type,Prop,Exists,Nonempty},
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[n]{/-}{-/},
  morestring=[b]",
}
\lstset{
  language=Lean,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{leankeyword}\bfseries,
  commentstyle=\color{leancomment}\itshape,
  stringstyle=\color{leanstring},
  breaklines=true,
  columns=flexible,
  showstringspaces=false,
  tabsize=2,
  frame=single,
  numbers=none,
  xleftmargin=1em,
  xrightmargin=1em,
}

\newcommand{\lean}[1]{\lstinline!#1!}
\newcommand{\N}{\mathbb{N}}

\title{Beyond Boolean Intersections: A Type-Safe Mechanism for Extracting\\Certified Shannon Bounds from D-Separated Traces}

\author{Anonymous Author(s)}



\begin{document}

\begin{abstract}
We give a type-safe, mechanized path from qualitative d-separation to certified
Shannon-style leakage bounds.  The path has four links:
d-separation, conditional independence, conditional data processing, and a
cut-set bound.  This paper mechanizes the graph-semantic front end in Lean~4;
a separate anonymized artifact mechanizes the finite quantitative
information-flow chain and is cited only as an anonymous artifact
\cite{companionqif}.

The graph-semantic core formalizes the two standard d-separation
characterizations---trail blocking~\cite{pearl2009causality} and moralized
ancestral-graph separation~\cite{lauritzen1996graphical}---as a verified
bisimulation between an operational semantics of active trails and a static
analysis of moralized-graph reachability.  Its three layers are a type-safe
graph syntax whose pairwise-disjointness invariant
($\text{DisjointSets}\,X\,Y\,Z$) acts as an affine ownership constraint; a
type-indexed intermediate representation ($\text{MAGWalk}$,
$\text{StaticRoute}$, $\text{OpenTrace}$, $\text{ActiveRoute}$); and
constructive witness-extraction pipelines in both directions.

The forward direction is a certified trace optimizer: it compiles any active
trail through a Bayes-ball state machine into a compressed moral-graph walk,
certifying soundness ($\text{moral-graph separation} \Rightarrow
\text{trail blocking}$) under the exact domain restriction that repairs
equivalence.  The reverse direction is a constructive decompiler of semantic
flow: from moral-graph reachability we reconstruct an explicit active-trail
witness, including a proved bad-collider normalization that strictly decreases
a well-founded measure.  Both directions compose into a closed equivalence on
the pairwise-disjoint domain (\lean{dSeparated_iff_dSeparates}); the graph
core carries no proof debt.  The local \lean{InfoTheoryBridge.lean} file
contains two intentional port placeholders for the artifact-proved semantic
bridge, and the remaining open items are isolated in Technical Debts.
\end{abstract}
\maketitle

\section{Introduction}\label{sec:intro}

The goal is not merely to decide whether two nodes are separated in a graph.
The larger target is a certified route from a graph query to a quantitative
information-flow statement:
\[
\text{d-separation} \Rightarrow \text{conditional independence}
\Rightarrow \text{conditional DPI} \Rightarrow \text{cut-set Shannon bound}.
\]
Across this Lean artifact and a separate anonymized artifact
\cite{companionqif}, that route is mechanized: the present development proves
the d-separation equivalence and witness machinery, while the quantitative
artifact proves the semantic d-separation-to-CI bridge, conditional Markov
instances, cut-set bounds, capacity certificates, and linear-chain DAG
instances.  The local bridge file in this repository is a port placeholder for
that proved chain, not the scientific boundary of the result.

The graph front end still has a sharp technical problem.  Consider the
simplest possible directed acyclic graph: a chain of three nodes
$0 \to 1 \to 2$.  Let $X = \{0\}$, $Y = \{1\}$, and $Z = \{0\}$.  Does $Z$
d-separate $X$ from $Y$?

The two standard textbooks give different answers:
\begin{itemize}
  \item \textbf{Trail blocking}~\cite{pearl2009causality} says: examine
    every undirected trail from $X$ to $Y$; if every internal triple is blocked by
    $Z$, the sets are d-separated.  The one-edge trail $0 \to 1$ has no internal
    triple, so it is not blocked.  Hence $X$ and $Y$ are \emph{not} d-separated.
  \item \textbf{Moral graph separation}~\cite{lauritzen1996graphical} says: take the ancestral
    subgraph of $X \cup Y \cup Z$, moralize it, delete $Z$, and check whether $X$
    and $Y$ are disconnected.  Node $0$ is in $Z$, so it is deleted.  The remaining
    graph has no path from $0$ to $1$, so $X$ and $Y$ \emph{are} d-separated.
\end{itemize}

We mechanize this discrepancy and its repair, treating the two
characterizations not as interchangeable definitions but as a \emph{verified
bisimulation} between two semantic layers:
\begin{itemize}
  \item an \textbf{operational semantics} of active trails (the ``execution''
    layer of information flow);
  \item a \textbf{static analysis} of moralized-graph reachability (the
    ``compiled'' layer).
\end{itemize}
The forward direction (soundness) is a \emph{certified trace optimizer}
that compiles an operational trace into static reachability evidence.  The
reverse direction (completeness) is a \emph{constructive decompiler of semantic
flow}: it reconstructs a concrete execution witness---an exploitability
certificate for the information leak---from static reachability alone.  The
counterexample says exactly that the two languages are \emph{not} bisimilar
without a well-typed domain restriction.

\subsection{Contributions}

Our contributions are:
\begin{enumerate}
  \item We formalize both d-separation characterizations in Lean~4 over finite
    DAGs with natural-number nodes, organized as a three-layer stack:
    type-safe graph syntax, certified intermediate representations, and
    constructive witness-extraction pipelines.
  \item We construct and machine-check a concrete counterexample, proving that
    unrestricted equivalence is false and identifying the exact repair
    ($\text{DisjointSets}\,X\,Y\,Z$).
  \item We mechanize the \textbf{forward bisimulation} (soundness): an explicit
    compilation pipeline
    $\text{Trail} \to \text{BayesBallPath} \to \text{MAGWalk} \to \allowbreak \text{Reachable}$,
    certified by node-survival proofs.
  \item We mechanize the \textbf{reverse bisimulation}
    (witness-producing completeness): a decompilation pipeline
    $\text{Reachable} \to \text{StaticRoute} \to \text{OpenTrace} \to \allowbreak \text{ActiveRoute} \to \exists\,\text{Trail},\;\neg\text{isBlocked}$.
    Every stage is type-indexed and constructive, including a proved
    bad-collider normalization (\lean{exists_split}, \lean{ancestor_escape},
    \lean{route_improves_of_bad}) that strictly decreases a well-founded
    measure.  The reverse pipeline carries no \lean{sorry}.
  \item We close the \textbf{full equivalence}
    (\lean{dSeparated_iff_dSeparates}): on the pairwise-disjoint domain,
    moral-graph separation holds if and only if every trail is blocked, with
    both directions discharged by the optimizer and the decompiler.
  \item We connect the graph core to a proved \textbf{quantitative
    information-flow} chain in a separate anonymized artifact
    \cite{companionqif}.  That artifact proves the semantic bridge
    \lean{factorizes_dsep_implies_cond_indep}, cut-set theorem
    \lean{abstract_cut_set_bound}, KKT capacity certificate
    \lean{capacity_le_of_kkt}, and end-to-end DAG-routed instance
    \lean{linear_chain_cut_set_bound_from_dag}.  The local
    \lean{InfoTheoryBridge.lean} file contains two intentional port
    placeholders for this artifact-proved bridge.
\end{enumerate}

The graph results are available as a standalone Lean~4 package.  The
decompiler produces not a mere existence proof but a computational
witness---a $\Sigma$-type containing the actual trail and the proof that it is
unblocked.

\section{A Three-Layer Architecture}\label{sec:architecture}

Our formal development is organized in three layers that mirror the architecture
of a verified compiler.

\subsection{Layer 1: Type-Safe Graph Syntax}

Nodes are natural numbers, edges are a finite set of ordered pairs
($\mathrm{Finset}\,(\N \times \N)$), and acyclicity is well-foundedness of the
edge relation.  The key safety invariant is
\lean{DisjointSets X Y Z}, which requires pairwise disjointness of the three
node sets.  In the language of substructural types, this is an \emph{affine
ownership constraint}: the conditioned variables $Z$ are borrowed by the query
and cannot alias the source $X$ or sink $Y$.  The intrinsic \lean{DAG} above
is the \emph{currently verified} syntax.  An extrinsic first-order surface
calculus that elaborates into it---the planned POPL-track front end---is
deferred to Section~\ref{sec:planned}; its current-vs-planned boundary is made
explicit in Table~\ref{tab:cvp} and Appendix~\ref{app:depplan}.

\begin{lstlisting}[caption={DAG structure and safety predicate}]
structure DAG where
  nodes : Finset \N
  edges : Finset (\N \times \N)
  edges_subset : \forall {u v}, (u,v) \in edges -> u \in nodes \wedge v \in nodes
  acyclic : WellFounded fun u v => (u,v) \in edges

def DisjointSets (X Y Z : Finset \N) : Prop :=
  Disjoint X Y \wedge Disjoint X Z \wedge Disjoint Y Z
\end{lstlisting}

\subsection{Layer 2: Certified Intermediate Representations}

Between operational trails and graph reachability sits a family of IRs that
preserve just enough structure for certified translation.

\paragraph{MAGWalk.} A compressed walk language equivalent to moral-graph
reachability.  It has three constructors: reflexivity, a single directed or
reversed edge (\lean{MAGWalk.single}), and a ``moral jump'' over an active
collider (\lean{MAGWalk.jump}).  The collider node is skipped, which is why
compression is correct.

\paragraph{StaticRoute.} A type-valued decompiler IR that keeps the witness
hidden by graph reachability.  Each step is either a forward edge, a backward
edge, or a moral jump storing the common child:

\begin{lstlisting}[caption={Static route IR (excerpt)}]
inductive StaticStep (G : DAG) (X Y Z : Finset \N) : \N -> \N -> Type where
  | directForward  (hEdge : G.HasEdge u v) ...
  | directBackward (hEdge : G.HasEdge v u) ...
  | moralJump (huw : G.HasEdge u w) (hvw : G.HasEdge v w)
      (hw : w \in G.ancestralSubgraphNodes (X \cup Y \cup Z)) ...
\end{lstlisting}

\paragraph{OpenTrace.} The compiler target before converting to an active route.
Each \lean{cons} node stores its oriented DAG traversal \emph{and} a local
non-blocking proof, making \lean{OpenTrace} a type-valued certificate that every
junction is $Z$-open.

\subsection{Layer 3: Witness-Extraction Pipelines}

The two pipelines are summarized in Figure~\ref{fig:pipeline}.

\begin{figure}[h]
\centering
\resizebox{\textwidth}{!}{
\begin{tikzpicture}[
  node distance=1.8cm and 2.4cm,
  every node/.style={align=center, font=\small},
  box/.style={draw, rounded corners, minimum width=2.2cm, minimum height=0.9cm, fill=blue!5},
  arrow/.style={-{Stealth}, thick},
  label/.style={font=\scriptsize\itshape, above, sloped}
]
  % Forward (optimizer)
  \node[box] (trail) {Active trail\\$t : \text{Trail}\,G\,x\,y$};
  \node[box, right=of trail] (bb) {Bayes-ball path\\$p : \text{BayesBallPath}$};
  \node[box, right=of bb] (mag) {MAGWalk\\$w : \text{MAGWalk}\,G\,X\,Y\,Z$};
  \node[box, right=of mag] (reach) {Graph reachability\\$\text{Reachable}_{G'}\,x\,y$};

  \draw[arrow] (trail) -- node[label]{lift} (bb);
  \draw[arrow] (bb) -- node[label]{compress} (mag);
  \draw[arrow] (mag) -- node[label]{embed} (reach);

  % Reverse (decompiler)
  \node[box, below=1.2cm of reach] (rreach) {$\text{Reachable}_{G'}\,x\,y$};
  \node[box, left=of rreach] (rstatic) {StaticRoute\\$r : \text{StaticRoute}$};
  \node[box, left=of rstatic] (ropen) {OpenTrace\\$o : \text{OpenTrace}$};
  \node[box, left=of ropen] (ractive) {ActiveRoute\\$a : \text{ActiveRoute}$};
  \node[box, left=of ractive] (rtrail) {$\exists\,\text{Trail},\;\neg\text{isBlocked}$};

  \draw[arrow, dashed] (rreach) -- node[label]{decompile} (rstatic);
  \draw[arrow, dashed] (rstatic) to [out=110, in=70, looseness=5] node[label]{normalize} (rstatic);
  \draw[arrow, dashed] (rstatic) -- node[label]{compile} (ropen);
  \draw[arrow, dashed] (ropen) -- node[label]{bridge} (ractive);
  \draw[arrow, dashed] (ractive) -- node[label]{realize} (rtrail);

  % Bracket labels
  \node[above=0.3cm of trail, font=\bfseries\small] {Forward: Certified Trace Optimizer};
  \node[below=0.3cm of rtrail, font=\bfseries\small] {Reverse: Decompilation of Semantic Flow};
\end{tikzpicture}
}
\caption{The two witness-extraction pipelines.  Solid arrows: forward
(soundness).  Dashed arrows: reverse (completeness).  Every arrow, including
the \emph{normalize} step, is a total constructive function in Lean; the
normalize step is a proved bad-collider minimization
(Section~\ref{sec:reverse}).}
\label{fig:pipeline}
\end{figure}

\subsection{Planned Surface Calculus (POPL Track)}\label{sec:planned}

\emph{This subsection describes planned POPL-track work.  None of the
surface-calculus statements below are mechanized in the current artifact; the
verified core is exactly the intrinsic development of
Sections~\ref{sec:counterexample}--\ref{sec:reverse}.}

The current artifact verifies d-separation directly over the intrinsic
\lean{DAG} structure: nodes and edges are raw \lean{Finset}s and acyclicity is
a \lean{WellFounded} proof obligation discharged at construction.  For the POPL
submission we plan an \emph{extrinsic} first-order surface calculus so that
queries are written in a checked source syntax rather than assembled as raw
finite sets.

\paragraph{Design (planned).} The surface calculus is first order with an
explicit environment $\Gamma$ of node and edge declarations and \emph{no}
higher-order binders.  This is a deliberate scoping choice: keeping binders out
of the surface language confines the metatheory to query reasoning rather than
substitution- and closure-heavy proofs.  Surface graphs carry a syntactic
\emph{rank annotation}; a local rank check certifies acyclicity without global
well-foundedness reasoning at the surface level, and a well-ranked surface
graph elaborates into an intrinsic \lean{DAG} through the existing
\lean{DAG.ofRank} constructor (the rank certificate discharges the
\lean{WellFounded} obligation).  The intrinsic target \lean{DAG.ofRank} already
exists; the planned additions are the surface AST, the rank checker, and the
elaborator.

\paragraph{Planned metatheory.} Two obligations are planned, \emph{not} yet
proved: (i) \emph{elaboration soundness}---every well-ranked surface graph
elaborates to a valid \lean{DAG}; and (ii) \emph{query conservativity}---on the
pairwise-disjoint domain, a surface-level d-separation query agrees with the
intrinsic decision certified by \lean{dSeparated_iff_dSeparates}.  Obligation
(ii) is what lets the verified equivalence of Section~\ref{sec:reverse} be
reused unchanged behind the surface syntax.

Table~\ref{tab:cvp} fixes the boundary between what is mechanized today and
what is planned for POPL; Appendix~\ref{app:depplan} sketches the planned
theorem-dependency structure.

\begin{table}[h]
\centering
\small
\begin{tabular}{p{0.31\linewidth}p{0.24\linewidth}p{0.35\linewidth}}
\toprule
\textbf{Asset} & \textbf{Status} & \textbf{Lean declaration / note} \\
\midrule
Intrinsic \lean{DAG} + \lean{DisjointSets} & Verified & current core; \lean{DAG.ofRank} \\
Endpoint counterexample & Verified & \lean{dsep_complete_endpoint_in_Z_counterexample} \\
Soundness (disjoint domain) & Verified & \lean{dSeparated_of_dSeparated_disjoint} \\
Bad-collider normalization & Verified & \lean{route_improves_of_bad} \\
Reverse witness decompiler & Verified & \lean{activeWitness_of_not_dSeparated} \\
Closed equivalence & Verified & \lean{dSeparated_iff_dSeparates} \\
Surface AST + rank checker + elaborator & Planned (POPL) & extrinsic front end; not mechanized \\
Elaboration soundness / query conservativity & Planned (POPL) & metatheory obligations above \\
$d$-sep $\Rightarrow$ conditional independence & Mechanized (anonymous artifact) & \lean{factorizes_dsep_implies_cond_indep} \cite{companionqif} \\
Conditional DPI / cut-set reuse & Mechanized (anonymous artifact, zero-\lean{sorry}) & \lean{abstract_cut_set_bound}; \lean{capacity_le_of_kkt}; \lean{linear_chain_cut_set_bound_from_dag} \cite{companionqif} \\
\lean{InfoTheoryBridge.lean} port & 2 \lean{sorry} placeholders (POPL) & local port of the proved semantic bridge \\
\bottomrule
\end{tabular}
\caption{Current-vs-planned boundary.  ``Verified'' rows are mechanized in
this Lean~4 artifact with no \lean{sorry}; anonymous-artifact rows are proved
in the cited anonymous artifact; the \lean{InfoTheoryBridge.lean} row is a
local port debt, not a missing quantitative theorem.}
\label{tab:cvp}
\end{table}

\section{Background: The Two Criteria}\label{sec:background}

\subsection{Trail Blocking (Operational Semantics)}

An undirected \emph{trail} from $x$ to $y$ in a DAG $G$ is a finite walk that may
follow edges forward or backward.  A consecutive triple $a \to b \to c$ or
$a \leftarrow b \to c$ is a \emph{non-collider}; $a \to b \leftarrow c$ is a
\emph{collider}.  A triple is \emph{blocked} by $Z$ if:
\begin{itemize}
  \item it is a non-collider whose middle node is in $Z$; or
  \item it is a collider whose middle node and all descendants are disjoint from $Z$.
\end{itemize}
A trail is \emph{blocked} by $Z$ if some internal triple is blocked.  Then $Z$
\emph{d-separates} $X$ from $Y$ in the trail sense if every trail from $X$ to $Y$
is blocked by $Z$.

\subsection{Moral Graph Separation (Static Analysis)}

The \emph{ancestral subgraph} of $S$ is the subgraph induced by all ancestors of
nodes in $S$.  Its \emph{moral graph} is obtained by (1)~forgetting edge
directions and (2)~connecting all co-parents (pairs of nodes with a common child).
The \emph{d-separation graph} is the moral graph of the ancestral subgraph of
$X \cup Y \cup Z$ with $Z$ removed.  Then $Z$ d-separates $X$ from $Y$ in the
graph sense if no node of $X$ is connected to any node of $Y$ in this graph.

\subsection{Textbook Equivalence and Its Caveat}

Standard references state that the two criteria are equivalent under the implicit
assumption that $X$, $Y$, and $Z$ are pairwise disjoint.  In applications,
however, the disjointness precondition is often omitted.  Our counterexample
(Section~\ref{sec:counterexample}) shows that this omission is not benign.

\section{The Endpoint Caveat: A Counterexample}\label{sec:counterexample}

Our counterexample lives on the simplest non-trivial DAG:

\begin{center}
\begin{tikzpicture}[node distance=1.5cm, every node/.style={circle,draw,minimum size=0.7cm}]
  \node (n0) {$0$};
  \node (n1) [right of=n0] {$1$};
  \node (n2) [right of=n1] {$2$};
  \draw[-{Stealth}] (n0) -- (n1);
  \draw[-{Stealth}] (n1) -- (n2);
\end{tikzpicture}
\end{center}

Let $X = \{0\}$, $Y = \{1\}$, $Z = \{0\}$.

\begin{theorem}[Endpoint-in-$Z$ counterexample]\label{thm:counterexample}
There exists a DAG $G$ and finite sets $X, Y, Z \subseteq \N$ such that
$G$ d-separates $X$ from $Y$ in the moral-graph sense, but $Z$ does \emph{not}
d-separate $X$ from $Y$ in the trail-blocking sense.
\end{theorem}

\begin{lstlisting}[caption={Mechanized counterexample}]
theorem dsep_complete_endpoint_in_Z_counterexample :
    \exists (G : DAG) (X Y Z : Finset \N),
      DAG.dSeparated G X Y Z \wedge \neg dSeparates G X Y Z := by
  refine <chain3, {0}, {1}, {0}, ?_, ?_>
  ...
\end{lstlisting}

\paragraph{Why the disagreement happens.} Trail blocking only inspects
\emph{internal} triples.  The one-edge trail $0 \to 1$ has no internal triple at
all, so nothing can block it.  By contrast, the moral-graph construction deletes
$Z$ unconditionally; since $0 \in Z$, the edge $(0,1)$ disappears from the
moralized graph, and $0$ and $1$ are disconnected.

\paragraph{The standard repair.} If we require $X \cap Z = \emptyset$ and
$Y \cap Z = \emptyset$, the start node of any trail is never deleted, and the
endpoint caveat disappears.

\section{Forward Bisimulation: The Certified Trace Optimizer}\label{sec:forward}

Under $\text{DisjointSets}\,X\,Y\,Z$, we mechanize the soundness direction:
\[
\text{DAG.dSeparated}(G,X,Y,Z) \;\Longrightarrow\; \text{dSeparates}(G,X,Y,Z).
\]

The proof follows a three-stage compilation pipeline.

\paragraph{Stage 1: Trail $\to$ Bayes-ball path.} If a trail $t$ from $x \in X$
to $y \in Y$ is not blocked by $Z$, and $x \notin Z$ (guaranteed by
$\text{Disjoint}\,X\,Z$), then we lift $t$ to an explicit Bayes-ball state path.
The construction is by induction on the trail, analyzing each directional triple
to show that the Bayes-ball state machine can traverse it.

\begin{lstlisting}[caption={Trail to Bayes-ball path}]
def bayesBallPath_of_active_trail_outOf
    (t : Trail G u v) (h_active : \neg t.isBlocked Z)
    (huZ : u \notin Z) :
    \Sigma final_dir, BayesBallPath G Z (u, outOf) (v, final_dir)
\end{lstlisting}

\paragraph{Stage 2: Certified membership.} The Bayes-ball path contains
intermediate states that must ``survive'' deletion of $Z$.  We prove a certified
version that guarantees every \lean{RequiredState} node belongs to the
d-separation graph nodes:

\begin{lstlisting}[caption={Certified Bayes-ball path}]
def bayesBallPathCert_of_active_trail_outOf
    (t : Trail G u v) (h_active : \neg t.isBlocked Z)
    (huZ : u \notin Z)
    (huA : u \in G.ancestralSubgraphNodes (X \cup Y \cup Z))
    (hvD : v \in G.dSeparationGraphNodes X Y Z) :
    \Sigma final_dir, {p : BayesBallPath ... //
      \forall {q}, BayesBallPath.RequiredState p q ->
        q.1 \in G.dSeparationGraphNodes X Y Z}
\end{lstlisting}

\paragraph{Stage 3: Compress to MAGWalk.} A Bayes-ball path is a sequence of
single-edge steps.  The \lean{compress} algorithm scans it with a two-step
window: ordinary non-collider windows become a \lean{MAGWalk.single}, while
active-collider windows $(a,\_) \to (b,\text{into}) \to (c,\text{outOf})$ become
a single \lean{MAGWalk.jump} over the collider $b$.  The collider itself need not
survive deletion of $Z$, which is why this compression is correct.

\begin{theorem}[Soundness]\label{thm:soundness}
For any DAG $G$ and pairwise-disjoint finite sets $X, Y, Z \subseteq \N$,
if $G$ d-separates $X$ from $Y$ in the moral-graph sense, then every trail
from $X$ to $Y$ is blocked by $Z$.
\end{theorem}

\begin{lstlisting}[caption={Top-level soundness theorem}]
theorem dSeparated_of_dSeparated_disjoint
    (hXYZ : DisjointSets X Y Z)
    (hsep : DAG.dSeparated G X Y Z) : dSeparates G X Y Z
\end{lstlisting}

\section{Reverse Bisimulation: Decompilation of Semantic Flow}\label{sec:reverse}

The reverse direction is the decompiler: from $\neg\text{DAG.dSeparated}$ it
constructively reconstructs an active-trail witness---an exploitability
certificate showing that the conditioning set $Z$ fails to block the flow.
The pipeline is:
\[
\begin{aligned}
\text{Reachable}
&\;\to\; \text{StaticRoute}
\;\to\; \text{NormalizedStaticRoute} \\
&\;\to\; \text{OpenTrace}
\;\to\; \text{ActiveRoute}
\;\to\; \exists\,\text{Trail},\;\neg\text{isBlocked}.
\end{aligned}
\]

\subsection{From Reachability to StaticRoute}

Graph reachability is a \lean{Prop}-valued relation that hides the structure of
each step.  We introduce \lean{StaticStep} and \lean{StaticRoute} as type-valued
evidence: a \lean{StaticRoute} is an explicit list of directed edges, reversed
edges, and moral jumps, each carrying the witness needed for later stages.

\begin{lstlisting}[caption={Reachability to static route}]
theorem nonemptyStaticRoute_of_dSeparationGraph_reachable
    (h : (G.dSeparationGraph X Y Z).Reachable u v) :
    Nonempty (StaticRoute G X Y Z u v)
\end{lstlisting}

\subsection{Normalization: Eliminating Bad Colliders}

A ``bad collider'' is a collider window whose middle node and all descendants are
disjoint from $Z$.  Such a collider is blocked, so it must be eliminated.  Normalization may change endpoints: if a bad collider leaks
ancestry into $X$ or $Y$, we reroute the prefix or suffix to a different node in
that set.

We package static routes as endpoint-carrying witnesses
(\lean{StaticRouteWitness}) and define a bad-collider count \lean{routeBadCount}.
The minimization is a \lean{Nat.find} wrapper: if every nonzero-count witness can
be strictly improved, a zero-count witness exists.

\begin{lstlisting}[caption={Bad-collider minimization wrapper}]
theorem normalized_route_exists_of_improves
    (hwit : \exists w : StaticRouteWitness G X Y Z, True)
    (himprove : \forall w, routeBadCount w \neq 0 ->
      \exists w', routeBadCount w' < routeBadCount w) :
    \exists w, routeBadCount w = 0
\end{lstlisting}

The improvement obligation is discharged by the core normalization theorem,
which is fully proved (no \lean{sorry}):

\begin{lstlisting}[caption={Strict improvement for non-normalized routes}]
theorem route_improves_of_bad
    (w : StaticRouteWitness G X Y Z) (hbad : routeBadCount w \neq 0) :
    \exists w' : StaticRouteWitness G X Y Z,
      routeBadCount w' < routeBadCount w
\end{lstlisting}

The proof is a short dispatcher over three established lemmas.
\lean{exists_split} extracts the first bad collider as a \lean{Split} carrying a
zero-bad prefix and a count bound.  \lean{ancestor_escape} shows the bad child's
leaked ancestry must reach a node in $X$ or $Y$ (its descendant cone misses
$Z$, so it cannot dead-end).  \lean{bad_child_survives} and
\lean{escape_path_survives} prove that the rerouted escape chain stays inside
the d-separation graph nodes.  Reattaching the escape chain---backward into $X$
or forward into $Y$---removes a moral jump of cost $1$ and adds chains of cost
$0$; the strict decrease then follows by
\texttt{countBadColliders\_backwardEscape\_}\allowbreak\texttt{append\_suffix\_lt} (resp.\
\texttt{countBadColliders\_prefix\_append\_}\allowbreak\texttt{forwardEscape\_lt}) and \lean{omega}.  As a
deliberate design choice, the witness endpoints may change: rerouting is allowed
to land on a \emph{different} node of $X$ or $Y$, which is what makes the escape
construction total.

\subsection{From Zero-Bad StaticRoute to OpenTrace to ActiveRoute}

A static route with zero bad colliders compiles directly to an \lean{OpenTrace},
because every junction is locally open:

\begin{lstlisting}[caption={Zero-bad compiler}]
theorem openTrace_of_countBadColliders_zero
    (hzero : countBadColliders arrival route = 0) :
    Nonempty (\Sigma finalDir, OpenTrace G Z (x, arrival) (y, finalDir))
\end{lstlisting}

The \lean{OpenTrace} is then converted to a \lean{BayesBallPath} and wrapped as
an \lean{ActiveRoute}.  Finally, \lean{ActiveRoute.to_activeTrail} yields the
existential witness:

\begin{lstlisting}[caption={Final decompiler theorem}]
theorem activeWitness_of_not_dSeparated
    (hnot : \neg DAG.dSeparated G X Y Z) : ActiveWitness G X Y Z
\end{lstlisting}

The result is a constructive $\Sigma$-type containing the actual nodes $x \in X$,
$y \in Y$, the final direction, and the active route.

\subsection{The Closed Equivalence}

Composing the optimizer (Section~\ref{sec:forward}) and the decompiler yields
the full equivalence on the well-typed domain---the bisimulation
statement: on pairwise-disjoint $X,Y,Z$, the operational and static languages
agree exactly.

\begin{lstlisting}[caption={Full d-separation equivalence (no proof debt)}]
theorem dSeparated_iff_dSeparates
    (hXYZ : DisjointSets X Y Z) :
    DAG.dSeparated G X Y Z <-> dSeparates G X Y Z
\end{lstlisting}

\begin{remark}[Verification status]\label{rem:status}
The full Lean development typechecks (\lean{lake build}).  The graph-semantic
core---both directions of \lean{dSeparated_iff_dSeparates} and the entire
reverse synthesis stack---contains no \lean{sorry}.  The only \lean{sorry}s in
the package are two \emph{intentional} scaffold stubs in
\lean{InfoTheoryBridge.lean} (Section~\ref{sec:related}), which are out of
scope for the equivalence claims.
\end{remark}

\section{The PL Dictionary: Reframing Causal Graphs}\label{sec:dictionary}

To situate this work for a POPL audience, we reframe traditional causal-graph
vocabulary in programming-languages terms:

\begin{table}[h]
\centering
\small
\begin{tabular}{ll}
\toprule
\textbf{Causal Inference} & \textbf{PL Framing} \\
\midrule
DAG & Dataflow AST / Dependency Graph Syntax \\
$\text{DisjointSets}\,X\,Y\,Z$ & Affine ownership constraint (no aliasing of $Z$) \\
Active Trail & Operational trace of semantics \\
MAGWalk & Reachability abstract IR (compressed moral-graph walk) \\
Collider Jump (\lean{MAGWalk.jump}) & Peephole optimization / semantic fold \\
Trail $\to$ BayesBallPath & Typestate tracking \\
BayesBallPath $\to$ MAGWalk & Certified trace optimizer \\
StaticRoute $\to$ OpenTrace $\to$ ActiveRoute & Semantic-flow decompiler / exploitability-certificate generator \\
Cut-Set / Mutual Information & Quantitative Information Flow (QIF) capacity bounds \\
\bottomrule
\end{tabular}
\caption{Reframing causal-graph concepts for a POPL audience.}
\label{tab:dictionary}
\end{table}

\section{Related Work}
\label{sec:related}

Our mechanization sits at the intersection of causal inference, mechanized probabilistic semantics, and verified compilation. We briefly survey the most closely related lines of work.

\paragraph{Mechanized Causal Inference and D-Separation.}
Formalizing the graph-theoretic foundations of causal inference is an active frontier. Recently, Chancelier, De Lara, and Heymann~(2024) \cite{chancelier2024conditional} mechanized d-separation as a binary relation using Coq and MathComp/SSReflect, providing a relational-algebraic treatment of graphical separation. Parallel to our work, another Coq formalization \cite{coqci2025} formalized the semantics of conditional independence and d-separation, bridging structural properties with function-based definitions for experimental design.

Both developments adopt standard mathematical formulations without exposing the implicit textbook domain caveats. Our work contributes the first formalized counterexample and repair (the endpoint caveat) for the gap between syntactic trails and moralized node-deletion. We also shift from purely relational verification to a constructive bisimulation that generates executable IRs (e.g., \lean{OpenTrace}), optimized in Lean~4 for computational witness synthesis.

\paragraph{Categorical and Topological D-Separation.}
Fritz and Klingler~(2023) \cite{fritz2023dseparation}, extending work by Fritz and Liang~(2023) \cite{fritz2023free}, abstracted d-separation using string diagrams in Markov categories. Their framework bypasses moralization, reading conditional independence directly from topological connectedness in a generalized measure-theoretic space. Our approach is complementary: Markov categories provide a general topological theory, while our artifact targets the legacy operational graph definitions (moral graphs and trace trails) implemented in existing software. By defining \lean{DisjointSets} as an affine ownership type, we show that classical d-separation behaves like a resource-restricted syntax within a broader categorical structure.

\paragraph{Verified Trace Bisimulation and Witness Extraction.}
The forward pipeline (soundness) translates unbounded trace semantics into a compressed step-relation (\lean{MAGWalk}), analogous to the semantics-preserving translations in Leroy's CompCert~(2009) \cite{leroy2009formal}. The reverse pipeline (completeness) deterministically unwinds reachability into a certified active trail, resembling Necula's Proof-Carrying Code~(1997) \cite{necula1997proof}: instead of returning a boolean assertion of an information leak, our decompiler synthesizes an explicit, machine-checkable exploitability certificate. This provides a graph-theoretic substrate for future verified static analyzers of quantitative information flow.

\subsection{Quantitative Information-Flow Bridge}

\paragraph{Certified Shannon bounds.} Security analysis has moved from
non-interference (``0 or 1'') to measuring leakage in bits.  The verified
bisimulation here supplies the graph-semantic front end for such bounds: a
closed d-separation equivalence licenses replacing a conditional-independence
premise by a checked graph property in a cut-set capacity argument.  The full
finite chain is
\[
\text{d-separation} \;\Rightarrow\; \text{conditional independence}
\;\Rightarrow\; \text{conditional DPI} \;\Rightarrow\; \text{cut-set bound}.
\]

A separate anonymized artifact proves the quantitative half of this chain
\cite{companionqif}.  Its semantic bridge
\lean{factorizes_dsep_implies_cond_indep} proves d-separation implies
conditional independence under a stated \lean{FactorizesOverDAG} hypothesis;
the structural Verma--Pearl derivation of that hypothesis from a DAG
factorization remains a technical debt, not a gap in the semantic bridge
theorem.  The same artifact proves
\lean{condMarkov_of_factorizes_dsep_fourVar},
\lean{abstract_cut_set_bound}, \lean{capacity_le_of_kkt}, and the end-to-end
linear-chain result \lean{linear_chain_cut_set_bound_from_dag}.  The capacity
premise \lean{h_cap} in \lean{abstract_cut_set_bound} is a theorem
hypothesis: the theorem derives the cut-set bound from the stated capacity
upper bound, and \lean{capacity_le_of_kkt} supplies a mechanized route for
certifying such bounds from KKT witnesses.  The Gaussian-KL steps are carried
as explicit external premises (\lean{AssumesFanoBound},
\lean{AssumesMissedCellBound}), not hidden proof obligations.

In this repository, \lean{InfoTheoryBridge.lean} restates the first bridge in
the local DAG namespace and contains exactly two intentional \lean{sorry}s:
\lean{ConditionalIndependence} and
\texttt{dSeparation\_implies\_}\allowbreak\texttt{conditional\_independence}.
These are port
placeholders for the artifact-proved bridge.  The genuinely open items are
limited to the three debts listed in Appendix~\ref{app:depplan}.

\section{Conclusion}\label{sec:conclusion}

We formalized the two standard d-separation criteria in Lean~4 as a verified
bisimulation between an operational semantics of active trails and a static
analysis of moralized-graph reachability.  We machine-checked a concrete
counterexample to unrestricted equivalence, characterized the repair
(pairwise disjointness as an affine ownership constraint), and discharged both
directions: a certified trace optimizer for soundness and a constructive,
witness-producing decompiler---with a proved bad-collider normalization---for
completeness.  The two compose into the closed theorem
\lean{dSeparated_iff_dSeparates}, and the graph-semantic core carries no proof
debt.  Together with the cited anonymous artifact, this gives a mechanized path
from d-separated traces to certified Shannon cut-set bounds; the remaining
work is the narrow technical debt listed below.

\appendix

\section{Technical Debts}\label{app:depplan}

This appendix separates proved theorem chains from remaining work.
Figure~\ref{fig:depplan} sketches the dependency structure.  Solid nodes and
edges are mechanized with no project \lean{sorry} in this artifact or in the
cited anonymous artifact; dashed nodes are the only open or port items.

\begin{enumerate}
  \item \textbf{Structural Verma--Pearl derivation.}  The anonymous artifact
    proves the semantic bridge
    \lean{factorizes_dsep_implies_cond_indep}: under
    \lean{FactorizesOverDAG}, d-separation implies conditional independence.
    What remains open is the structural derivation of
    \lean{FactorizesOverDAG} from a standard DAG factorization theorem.
  \item \textbf{General capacity metatheorem.}  The anonymous artifact proves
    concrete capacity certificates, including \lean{capacity_le_of_kkt}, and
    uses them in cut-set bounds.  A general Blahut--Arimoto/KKT concavity
    sufficiency theorem remains future work.
  \item \textbf{Surface calculus and certified decision procedure.}  The
    current artifact verifies the intrinsic \lean{DAG} core.  The planned
    first-order surface syntax, rank checker, elaboration soundness theorem,
    query conservativity theorem, and certified d-separation decision
    procedure are not yet mechanized.
\end{enumerate}

Engineering note: the graph artifact and the quantitative artifact are two
physically separate Lake projects with near-isomorphic DAG definitions and no
shared imports yet.  The two \lean{sorry}s in this repository's
\lean{InfoTheoryBridge.lean} file are a port placeholder for the proved
semantic bridge in the cited anonymous artifact.

\begin{figure}[h]
\centering
\resizebox{\textwidth}{!}{
\begin{tikzpicture}[
  node distance=1.1cm and 1.4cm,
  every node/.style={align=center, font=\scriptsize},
  done/.style={draw, rounded corners, minimum height=0.8cm, fill=blue!5},
  plan/.style={draw, dashed, rounded corners, minimum height=0.8cm, fill=gray!8},
  sa/.style={-{Stealth}, thick},
  pa/.style={-{Stealth}, thick, dashed}
]
  \node[done] (cex) {\texttt{dsep\_complete\_}\\\texttt{endpoint\_in\_Z\_}\\\texttt{counterexample}};
  \node[done, below=of cex] (snd) {\texttt{dSeparated\_of\_}\\\texttt{dSeparated\_disjoint}\\(soundness)};
  \node[done, below=of snd] (dec) {\texttt{activeWitness\_}\\\texttt{of\_not\_dSeparated}\\(decompiler)};
  \node[done, right=of snd] (equiv) {\texttt{dSeparated\_}\\\texttt{iff\_dSeparates}};
  \node[done, right=of equiv] (ci) {\texttt{factorizes\_dsep\_}\\\texttt{implies\_cond\_indep}\\(proved)};
  \node[done, right=of ci] (cut) {\lean{abstract_cut_set_bound}\\+\lean{capacity_le_of_kkt}\\(proved)};
  \node[done, below=of cut] (chain) {\texttt{linear\_chain\_}\\\texttt{cut\_set\_bound\_from\_dag}\\(proved)};
  \node[plan, left=of snd] (surf) {surface calculus\\+ decision procedure\\(open)};
  \node[plan, above=of ci] (vp) {Verma--Pearl\\structural derivation\\(open)};
  \node[plan, above=of cut] (cap) {general capacity\\metatheorem\\(open)};
  \node[plan, below=of ci] (port) {\lean{InfoTheoryBridge.lean}\\port\\(2 \lean{sorry})};

  \draw[sa] (snd) -- (equiv);
  \draw[sa] (dec) -- (equiv);
  \draw[pa] (surf) -- node[font=\tiny, above, sloped]{elaborate} (snd);
  \draw[pa] (surf) to[bend right=20] node[font=\tiny, below, sloped]{conservativity} (equiv);
  \draw[sa] (equiv) -- node[font=\tiny, above, sloped]{feeds} (ci);
  \draw[sa] (ci) -- node[font=\tiny, above, sloped]{DPI} (cut);
  \draw[sa] (cut) -- (chain);
  \draw[pa] (vp) -- node[font=\tiny, left]{discharges hypothesis} (ci);
  \draw[pa] (cap) -- node[font=\tiny, right]{generalizes} (cut);
  \draw[pa] (ci) -- node[font=\tiny, right]{port} (port);
\end{tikzpicture}
}
\caption{Technical debts and theorem-dependency structure.  Solid: mechanized
in Lean~4 with no project \lean{sorry}.  Dashed: the three open debts and the
local \lean{InfoTheoryBridge.lean} port placeholder.}
\label{fig:depplan}
\end{figure}

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
